---
title: "DANA4810 Project - Part2"
output: html_document
date: "2026-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Github over Google docs for documenatiton
Imagine you are delivering a final presentation for the company with your teammates, and everybody is using Google Docs or Word, sending emails back and forth after each change. It is the final day, and people rushing to finish on time are overwriting each other's work, with no way to know who changed what. In the end, you lose time, nerves, and the quality of the work suffers as well.

GitHub solves these problems by giving every team member their own local copy of the files, tracking every change with a timestamp and author name, and allowing the team to merge contributions without overwriting each other's work.

## Getting started

We need to set up three thing before working on our documents:
1.Install Git<br>
2.Create a Github Account<br>
3.Connect RStudio to Github<br>

```{r}

```

## Basic Operation for Daily Use

1.Open RStudio<br>
2.Select a Project<br>
3.Pull a file that you would like to work on<br>
4.Commit your change<br>
5.Push to a Github<br>

```{r}

```

## Version Control

1.How to make a version<br>
2.How to check a history and differences<br>
3.How to restore from a staging area<br>
4.Hot to restore from a Github<br>
5.How to communicate with team members<br>

```{r}

```

## Tips

1.Do's & Don'ts<br>
2.Common Errors<br>

```{r}

```

## Document Update History

dd June 2026,  New Creation<br>

